---
title: '**Fabryka**'
subtitle: A web application for easily analysing & exploring the fabric of archaeological assemblages
output:
  pdf_document: default
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Fabryka}
- \fancyhead[R]{R web application for fabric analysis}
- \fancyfoot[C]{\thepage}
- \usepackage{float}
- \floatplacement{figure}{H}
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: console
bibliography: references.bib
link-citations: true
---

```{=html}
<style>
body {
text-align: justify}
</style>
```

\begin{figure} [H]
\centering
  \includegraphics[width=4cm]{figure/hex_Fabryka4.png}
\label{}
\end{figure}

Marc THOMAS ^1\*^ (https://orcid.org/0000-0002-8160-1910) \smallbreak

^1^ TRACES, UMR 5608 - Université Toulouse Jean Jaurès, 5 Allée Antonio
Machado, FR-31058 Toulouse Cedex 9 \smallbreak

^\*^ Corresponding author \smallbreak

E-mail: [marcthomas1\@hotmail.fr](mailto:marcthomas1@hotmail.fr){.email}
\bigbreak

**Key words** : R shiny web application, fabric, site formation
processes, spatial analyses, orientation and inclination \bigbreak

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract

This article presents Fabryka, an open access R shiny web application
dedicated to the fabric analysis of archaeological or geological
assemblages. The fabric is a statistic made of the orientation and
inclination of elongated sedimentary particles and archaeological
objects. It provides information about the source of sediments, the
nature of the geological processes involved in the formation of the
deposits and their impact on the preservation of the archaeological
assemblages. These statistical analyses are not easily accessible to
non-specialists and are particularly time-consuming, sometimes requiring
the use of several software, some of which are chargeable. Fabryka uses
classical analytical procedures (Benn, rose and Schmidt diagrams and
statistical tests on orientations) for analysing the fabrics of objects
and includes a spatial method. New statistical methods and new ways of
exploration are proposed in the application and this present article.
Also, a call for contributions is proposed to construct a new reference
model with spatial information. \smallbreak

You can use Fabryka $\beta$ 3.0 by following this link
<https://marchaeologist.shinyapps.io/fabryka/>. The version will be
updated to Fabryka 1.0 after the review process of this article. The
Fabryka application files, made in R [@rcoreteam2025] by using *shiny* package [@chang2024], and this article, written in rmarkdown [@allaire2014; @xie2018; @xie2020], are both available online in a public github repository (<https://github.com/marchaeologist/fabryka/tree/main>). When reviewing this article you can fork and clone the fabryka project repository on github to make suggestions for modifications.\pagebreak

# Introduction

Fabric is made of the orientation and inclination of elongated
sedimentary particles [@bertran1997; @bertran2002; @lenoble2004]. It is
a statistic (*i.e.*, it needs as much measurements as possible) that
enables the identification of sediment sources that make up the deposits
[@texier2020], sedimentary processes involved in their formation
[@bertran1997; @bertran2002; @lenoble2004] and their impact on
archaeological assemblages [@isaac1967; @schick1986; @thomas2019].
Objects abandoned by prehistoric people can be considered as sedimentary
particles and are affected by the same processes [@schiffer1983]. As
geological particles are often not elongated enough, the fabric analysis
made of the sedimentary deposits depends mainly on the archaeological
material (bones and lithics), which enables them to be inserted into the
formation dynamics of the site and to obtain information on their
representativeness for the reconstruction of past activities.

One of the first archaeological application probably dates to the
experiments carried out by the South African G. Isaac (1967) on the
shores of Lake Magadi in Kenya. G. Isaac wanted to explain the role
played by hydrological processes on the bone accumulations found in the
Great Rift Valley. In November 1964, he placed several concentrations of
experimental material near a river branch. A few months later, heavy
rains had reactivated the valley's rivers. In addition to the
displacement or even loss of all or part of the material, G. Isaac
revealed a reorientation that was mainly perpendicular and, to a lesser
extent, parallel to the direction of flow, with a slight upstream
inclination. Since G. Isaac's pioneering work, there has been an
abundant literature devoted to the analysis of fabrics [@benn1994;
@lenoble2004; @mcpherron2018] allowing this method to be increasingly
performed on archaeological sites. Several software allow data
processing, some of which are not freely accessible, do not allow
spatial exploration of the data or do not offer interfaces to facilitate
the task of non-programmer users [@mcpherron2018].

The following sections explain all the selected approaches and our
proposals for processing the fabrics thanks to fabryka $\beta$ 3.0, from
taking measurements in the field, through the statistical methods and
graphical modes of representation used in the application (fig.
\ref{fig:figure_presentation}). \bigbreak

\begin{figure} [h]
\centering
  \includegraphics[width=16cm]{figure/presentation_applications.png}
  \caption{Fabryka creation scheme and organisation of the article}
\label{fig:figure_presentation}
\end{figure}

# Field data acquisition

During excavation, the fabric of elongated remains is generally
systematically measured. The direction (= bearing) of the lowest point
is measured in the range [0; 360] with North as the reference (*i.e.*
direction = 0°). The inclination (= plunge, dip) is measured in the
range [0; 90] with horizontal as a reference. The direction [0; 360] can
be converted into an orientation measured in the range [0; 180] by
subtracting 180° from the measurements in the range ]180; 360].

Traditionally, the direction of the objects lowest point is measured
using a compass and the inclination is measured using an inclinometer.
The aim is to measure the direction and inclination of the physical axis
of elongation of objects ignoring its shape. To do this, the measuring
device must be placed parallel to the general axis of elongation of the
piece without touching it. Measuring the fabric from its imprint or
visible surface is biased by its shape, making impossible to assess its
true inclination. Since the theodolite was introduced to excavations in
the 1990s, the fabrics is sometimes measured by recording the
coordinates of the ends of objects. In the latter case, the rules of
trigonometry are applied to calculate directions and inclinations.

Recently, speleologist have created a digital inclinometer (Leica
DistoX310 modified to DistoX2; <https://paperless.bheeb.ch/>) which can
be used to measure fabrics [@discamps2024]. The device stores and
associates data with the order of measurement. When the device is tilted
downwards, the direction of the lowest point is measured and, the
recorded inclination is negative. When the device is tilted upwards, the
direction of the highest point is recorded, and the inclination is
positive. To obtain data that is suitable for statistical analysis, in
the first case the absolute value of the inclination must be retained.
In the second case (*i.e.* positive inclination), if the measured
direction is in the interval [0; 180], 180° must be added to the
direction, and if the measured direction is in the interval ]180; 360],
180° must be subtracted from the direction.

When possible, preferring the use of a digital inclinometer avoids many
errors during measurement and data recording. On the other hand, for the
reasons given above, the use of a theodolite to measure fabrics is not
recommended as it results in inaccurate inclination measurements. As
well, measuring orientations and inclinations in sections leads to
sampling biases.

A sample corresponds to all the direction and inclination measurements
taken within a Lithostratigraphic Unit (UL). UL seems to be the most
coherent entity which needs to be explored because they are defined
according to sedimentary criterion, which a priori guarantee a certain
coherence in terms of the formation dynamics we are seeking to approach.
In Fabryka, analyses are constrained within samples unless the user
decides otherwise in the spatial exploration panel. \bigbreak

# Mathematics

## Classical method

Two main statistical methods are used:

1.  The first considers the orientation only. The calculation of the
    orientation rate 'r' (or 'L' when expressed as a percentage,
    Curray's [-@curray1956] formula tests the hypothesis of a unimodal
    distribution of orientations. If the 'p' value returned by the
    Rayleigh test is less than a threshold of 0.05, the null hypothesis
    of a random orientation from a uniform distribution is rejected. The
    formula for calculating the orientation rate 'r' where '$\alpha$' is
    the orientation in radians and 'n' is the sample size is as follows:

Curray's formula to calculate 'L' is (**1.**)

$$L = \frac{r*100}{n}$$ with (**2.**)

$$r = \sqrt{\sum_{i=\alpha}^n sin{(2\alpha)} + \sum_{i=\alpha}^n cos{(2\alpha)}}$$
\smallbreak where 'n' is the number of measures and '$\alpha$', is the
orientation (0 to 180°) \smallbreak

The Rayleigh test is performed thanks to the formula (**3.**)

$$p = e^{(-(n*L^2))*10^-4}$$

The method consisting of doubling the value of the orientations before
calculating the orientation rate L and the Rayleigh test, as proposed by
Krumbein [-@krumbein1939], tests the hypothesis of a bimodal distribution
of orientations with a period of 90°. The hypothesis of a 90° period
bimodal distribution of a sample of orientations is retained when p \<
0.05 [@curray1956]. The sensitivity of the Rayleigh test to the sample
size 'n' prevents samples from being compared with each other when 'n'
varies [@curray1956; @bertran2002]. For example, when the orientation
rate "L" is 25%, the Rayleigh test gives p-values below a critical
threshold of 0.05 for series with more than 50 measurements (fig. 2).

2.  The second considers the direction and inclination of the objects.
    The normalised eigenvalues E1, E2 and E3 are calculated from the
    cosine formed by the axis of the objects with the three axes of
    space [@watson1966]. The Benn indices [@benn1994] describe the
    fabric shape of a sample. The elongation index EL is equal to
    1-(E2/E1) and the isotropy index IS is E3/E1. When EL is high, the
    axes of the objects are grouped around a direction. When IS is high,
    the fabric is said to be isotropic meaning the axes of the objects
    show no order. Finally, when the value 1-EL-IS is high, the fabric
    is said to be planar, with the axes of the objects grouped around a
    plane. \bigskip

```{r rayleigh, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Evolution of the p-value returned by the Rayleigh test as a function of the orientation rate 'L' and the sample size 'n' \\label{rayleigh}", fig.height = 3.5, fig.pos="f"}

require(ggplot2)
L = c(0:60)

rayleigh <- function(L,n){
  p = exp(-(n*L^2)*10^-4)
  p_value <- p
  p_value
}

p_value_sim_30 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_30[i]<- rayleigh(i,30)
}

p_value_sim_40 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_40[i]<- rayleigh(i,40)
}

p_value_sim_50 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_50[i]<- rayleigh(i,50)
}

p_value_sim_100 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_100[i]<- rayleigh(i,100)
}

seuil_0.05 = c(rep(0.05, length(L)))

rayleigh_table <- data.frame(L, p_value_sim_30, p_value_sim_40, p_value_sim_50, p_value_sim_100,seuil_0.05)
  
ggplot(rayleigh_table, aes(x=L)) + 
    geom_line(aes(y = p_value_sim_100, colour = "n = 100"), size = 1) +
    geom_line(aes(y = p_value_sim_30, colour = "n =  30"), size= 1) +
    geom_line(aes(y = p_value_sim_40, colour = "n =  40"), size= 1) + 
    geom_line(aes(y = p_value_sim_50, colour = "n =  50"), size= 1) +
    geom_line(aes(y = seuil_0.05, colour = "threshold 0.05"), size= 0.5) +
    xlab("orientation rate 'L' (%)") +
    ylab("p-value") + 
    scale_color_manual(values=c("#0073C2", "#EFC000", "#868686", "#3B3B3B", "red")) +         scale_x_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60)) +
scale_y_continuous(breaks = c(0, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.70, 0.80, 0.90, 1)) +
  theme_minimal() +
  theme(legend.position = "bottom") + 
  labs(colour = "")


```

## Spatialised method

To distinguish different statistical populations, a spatial method
proposed by McPherron (2018) is used. Thanks to this methods it is
possible to calculate the Benn indices and to test the orientation rate
'L' (formula (**3.**) modified R [@mcpherron2018] script) for each
remains, and its n nearest neighbours in 3 dimensions (McPherron
[-@mcpherron2018] R script, modified , formula (**4.**)). The analysis
of the fabric as proposed by McPherron [-@mcpherron2018] has the
advantage of spatialising the measurements and thus avoiding the
arbitrary sampling classically carried out by square and décapage. It
is possible to localize the statistical signal of fabrics by varying the
number of measurements sampled to calculate the Benn indices. The choice
of about 40 to 60 measurements (each remain and its 40 to 60 nearest
neighbours) to calculate the Benn indices allows local particularities
to be observed, and above all, it allows the Rayleigh test to be carried
out in its field of application (stability of the test between around 40
to 50 measurements after [-@bertran2002]).

By avoiding the instability of the Rayleigh test, the rejection rate
'Tr' of the null hypothesis provides a more reliable description of the
distribution of orientations in a sample. In addition, it allows
comparisons between samples despite varying sample size.

The rejection ratio 'Tr' is performed thanks to the formula (**4.**)

$$Tr = \frac{(n~times~p.value < 0.05)* 100}{n}$$

# Graphical representation

Several methods of representation are used. Rose diagrams summaries the
orientation and inclination measurements for each sample. Benn indices
are projected in a Benn diagram and measures (directions and plunge) are
projected into Schmidt diagrams. These figures are available for both
'classical' and 'spatialised' methods.

As proposed by McPherron (2018), a spatial investigation of the fabric
is proposed in a dedicated panel. Benn indices (1994) calculated for
each spatial series of remains are projected within a Benn diagram in
which each pole (isotropic, linear, and planar) is coloured respectively
in red, green and blue. The RGB colour code is reused to materialize the
fabric of the remains on spatial projections (plan and section views) of the material along the axes of the excavation grid. \bigbreak

# Interface and user interactivity

The next sections provide a description and a tutorial of Fabryka.

## Upload data

The simplest method to upload data in the application is to name the column names directly, as shown in the example files. If you miss a context column (*i.e.* 'id', 'code' or 'sample'), please create it by repeating a same character. By doing so, you do not need to select your variables in the main panel. The application correctly reads the data file directly. 

First, upload your .csv file in the left sidebar panel (fig.
\ref{fig:figure_upload_your_data}A). Pay attention to your decimal and
column delimiters (fig. \ref{fig:figure_upload_your_data}B and C).

Fabryka supports three angle measurement methods:

-   angles taken with a compass and an inclinometer
-   angles taken with a DistoX2
-   angles taken with two shots at the total station.

Five cases (*i.e.* forms of data) are covered by the application (fig.
\ref{fig:figure_upload_your_data}D):

-   check 'Case 1: Only angles' if you have in your dataset the columns
    'id', 'sample', 'code', direction' and 'dip'.
-   check 'Case 2: Only angles from DistoX2' if you have in your dataset
    the same columns that above but your data comes from a DistoX2 and
    needs to be converted.
-   check 'Case 3: Angles from DistoX2 with coordinates' if you have in
    your dataset the same columns that above (Case 2, data from DistoX2)
    with a single set of coordinates (x, y, z).
-   check 'Case 4: Angles and coordinates' if your data has the same
    form that above (Case 3) but is coming from a compass and a
    clinometer.
-   check 'Case 5: Two shots data without angles' if you have the
    columns 'id', 'sample', 'code' and the angles have been measured
    with a total station (*i.e.* two shots data).

\begin{figure} [h]
\centering
  \includegraphics[width=16cm]{figure/upload_your_data.jpg}
  \caption{A. Browse your .csv file, B. choose your column separator, C. choose your decimal separator, D. check the right form of your dataset, E. to do so, have a look of example files, F. select the column needed in the application, G. a new dataset with new columns appears, your are ready to explore your data.}
\label{fig:figure_upload_your_data}
\end{figure}

According to your method of angle measurement and the presence of
coordinates you must select the right case. Please refer to the example
files in the left side bar panel (fig.
\ref{fig:figure_upload_your_data}E). It contains a dataset coming from
Le Moustier Lower shelter (Dordogne, France). The example files directly
contain the good column names. You can use them directly in the
application. Whatever the names of the columns in your file, you can
select the appropriate columns in the column selector (fig.
\ref{fig:figure_upload_your_data}F). Please keep only complete lines
(*i.e.* no empty modalities, so no line without a fabric measurement).
When a data table with new columns appears below the column selector,
you are ready to use the application.

If your data comes from a DistoX2, the columns 'direction' (= 'bearing',
0 to 360° angles) and 'dip' (= 'plunge') are automatically corrected in
the new table. Whatever the form of your dataset, two new columns
'orientation_pi' (0 to 180° angles) and 'angle_double' (see Krumbein
1939) appear. If the angles were measured using a total station (*i.e.*
two shots data), the bearing and the plunge are also calculated. If only
one set of coordinates is provided with angles, a second set of
coordinates is calculated and displayed in the new table (fig.
\ref{fig:figure_upload_your_data}G). You can download the new data by
clicking the 'copy', the 'csv' or the 'pdf' buttons. \bigbreak

## Classical method

The classical method (fig. \ref{fig:figure_classical_method}) can be
computed for all sources of data and absence/presence of coordinates
(*i.e.* all cases described above).

This panel includes:

-   Benn diagrams with possible integration of Bertran and Lenoble (2002) model (fig. \ref{fig:figure_classical_method}D)
-   Rose diagrams for bearing and plunge (fig. \ref{fig:figure_classical_method}G)
-   Schmidt diagrams (fig. \ref{fig:figure_classical_method}H)
-   Summary data table with Benn indices, additional statistics, and orientation statistical tests.

In this last table (fig. \ref{fig:figure_classical_method}I):

-   'n' stands for number of measures in the samples
-   'E1', 'E2' and 'E3' are the eigenvalues (see Woodcock 1977). They
    correspond to the sum of the cosines formed by the axis of the
    objects and the three axis of the space. They are used to compute
    the Benn indices
-   'IS', 'PL' and 'EL' are calculated thanks to the eigenvalues (Benn
    1994). These are the Benn indices, ratios standing respectively for,
    'isotropy' with IS = E3/E1, 'elongation' with EL = 1-(E2/E1) and
    'planar' PL = 1-IS-EL
-   'L' is the vector magnitude (strength of the preferred artefact
    orientation) and 'R.p' is the p-value result of the Rayleigh test.
-   'L.double' and 'R.p.double' are the same statistics that above for
    double angles (Krumbein 1939).

You can control the data you plot and the data appearing in the figures
and the summary data table by changing the checkboxes filters ('filter by sample'
and 'filter by code') and the slider (*i.e.* minimum number of measures) in the left
sidebar panel (fig. \ref{fig:figure_classical_method}A, B and C).
Certain geological processes can have a different impact on objects
depending on their physical properties (*e.g.* silex versus bones in a
water flow as in the example files provided where bones have a strong
preferential orientation and are much more 'linear' in the Benn diagram,
see [@thomas2019] and [@texier2020]). The natural processes models that
you can add to the Benn diagram are calculated using data mainly from
[@bertran1997; @bertran2002; @bertran2006; @lenoble2004]. You will find the reference model data (with corresponding litarature references) which is used in the application in the 'model reference data' subpanel. In the rose diagram, orientations are considered (and not directions) as the Rayleigh test is perform on orientations (Curray 1956) in the
summary table. You can download all the figures by selecting a file format (.pdf, .jpeg and .png) and pressing the 'download' button (fig. \ref{fig:figure_classical_method}E and F).

\begin{figure} [h]
\centering
  \includegraphics[width=17cm]{figure/classical_method.jpg}
  \caption{A. filter by sample, B. if you want to filter by codes, uncheck 'no code distinction', C. manage the minimum sample size to perform analyses, D. buld your model thanks to Bertran and Lenoble (2002) data, E. and F. download figures by selecting a file format and pressing the 'download' button, G. rose diagrams subpanel, H. Schmidt diagram subpanel, I. summary table with additional statistics}
\label{fig:figure_classical_method}
\end{figure}

## Spatialised method

This panel includes (fig. \ref{fig:figure_spatialised_method}):

-   Benn diagrams with Benn indices computed with nearest neighbours points (McPherron 2018)
-   Spatial projections of the points according to their fabric shape
-   Summary data tables with orientation statistical tests, one at the
    scale of each spatial series (each object and its n nearest
    neighbours fabric = one row) and one at the scale of the sample (fig.
    \ref{fig:figure_spatialised_method}F and G).

In these tables:

-   'n' stands for number of measures in the samples
-   'E1', 'E2' and 'E3' are the eigenvalues [@woodcock1977]. They
    correspond to the sum of the cosines formed by the axis of the
    objects and the three axes of the space. They are used to compute
    the Benn indices.
-   'IS', 'PL' and 'EL' are calculated thanks to the eigenvalues (Benn
    1994). These are the Benn indices, ratios standing respectively for,
    'isotropy' with IS = E3/E1, 'elongation' with EL = 1-(E2/E1) and
    'planar' PL = 1-IS-EL
-   'L' is the vector magnitude (strength of the preferred artefact
    orientation) and 'R.p' is the p-value result of the Rayleigh test.
-   'L.double' and 'R.p.double' are the same statistics that above for
    double angles (see Krumbein 1939).

At the spatial series scale (*i.e.* each points and n neighbours),
statistics are provided in a table (Spatial series data table subpanel)
with the Benn indices, the orientation rates L, L.double and p-values
returned by the Rayleigh tests. At the sample scale, mean and standard
deviations values of Benn indices, mean L and its corresponding Rayleigh
tests means (mean values of all spatial series in the sample) and
rejections ratios 'Tr' (formula **4.**) are given in the subpanel
'Summary table of samples'.

In the projections, the orientation of the objects can be materialized
by sticks (by unchecking the corresponding box in the left sidebar
panel, fig. \ref{fig:figure_spatialised_method}D) which colour is
representative of the fabric of the series formed by the object and its
n nearest neighbours. You can also add orthophoto in the background of
projections (fig. \ref{fig:figure_spatialised_method}E). In the left
sidebar panel, you can control the samples considered in the analyses by
selecting the sample explored or by distinguishing the code used (*e.g.*
flint versus bone, fig. \ref{fig:figure_spatialised_method}A and B). You
can also use the slider to change the number of neighbors searched for
around each point (fig. \ref{fig:figure_spatialised_method}C). Be
careful, this will alter the results obtained in the Benn diagram and
Rayleigh tests. \bigbreak

## Spatial exploration

This panel includes (fig. \ref{fig:figure_spatial_exploration}):

-   A spatial exploration of your data starting from a Benn diagram
-   A spatial exploration of your data starting from the spatial
    projections (fig. \ref{fig:figure_spatial_exploration}J)
-   Summary data tables with orientation statistical tests (fig.
    \ref{fig:figure_spatial_exploration}I and K).

\begin{figure} [h]
\centering
  \includegraphics[width=17cm]{figure/spatialised_method.jpg}
  \caption{A. filter by sample, B. if you want to filter by code, uncheck 'no code distinction', C. choose the n nearest neighbors to perform the statistics, D. make sticks in spatial projections, E. upload a georeferenced orthophoto and check 'include' orthophoto to display it, F. new dataset at the spatial series scale, G. summary table at the sample scale.}
\label{fig:figure_spatialised_method}
\end{figure}

In this panel, you can perform a spatialised analysis of your data
starting from a Benn diagram or from projections. In both cases, you
need to select the data you want to explore more locally in a new
analysis. From the projections, you can select the points you want to
explore directly. In the Benn diagram, you need to select the lasso
selector (fig. \ref{fig:figure_spatial_exploration}D). Pay attention to
the number of points selected and check that the new sample contains
enough points to carry out the new analysis. If this is not the case,
you can lower the slider in the left sidebar panel to perform the
analysis with fewer neighbours (fig. \ref{fig:figure_spatial_exploration}E). You can download the new samples you have selected with spatial statistics from the corresponding
subpanels (fig. \ref{fig:figure_spatial_exploration}I and K). \bigbreak

## Tutorial

The 'Tutorial' panel contains a summary of the mathematics, statistics
and methods used in Fabryka and a short tutorial directly accessible
from the application. \bigbreak

\begin{figure} [h]
\centering
  \includegraphics[width=17cm]{figure/spatial_exploration.jpg}
  \caption{A. if you want to filter by sample, uncheck 'No sample distinction' B. if you want to filter by code, uncheck 'no code distinction', C. choose the n nearest neighbors to perform the statistics (first Benn diagram), E. choose the n nearest neighbors to perform the statistics on the sampled data (second Benn diagram), F. make sticks in spatial projections, G. you can upload a georeferenced orthophoto and check 'include' orthophoto to display it, H. a new spatial exploration of the fabric is performed, I. new dataset at the selected spatial series scale, I. a similar spatial exploration is possible from the projections, K. new dataset at the selected spatial series scale.}
\label{fig:figure_spatial_exploration}
\end{figure}


## New models

The 'spatialised' method generates often more linear fabrics in
comparison with the 'classical' method. This can be explained by the
fact that objects that are close in space are more likely to resemble
each other than objects that are further in space.

Also, the structure of the point pattern is evocative of its fabric.
Thus, the dispersion of the points depends on the shape of the fabric.
Two extremes appear: the most isotropic samples are those with the most
dispersed point patterns (*i.e.* the standard deviations are high) while
the most linear samples are those with the least dispersed point
patterns. Planar samples highlight point patterns with intermediate
dispersion. The explanation of this phenomenon is intuitive: the more
similar the fabrics of objects are, the closer they are in the Benn
diagram and the more linear they are. To reflect this phenomenon, the
standard deviations of the Benn indices are calculated and aim to
describe the dispersion point patterns in the Benn diagrams.

Global planar or isotropic fabrics can sometimes mask local anisotropy.
Within certain samples, groups of objects form a dense cluster in the
point pattern. This is expressed either by a locally more linear fabric,
or by a planar fabric with a strong bimodal orientation.

For the reasons given above, we wish to create new models that
considered the particularities of the spatial method. A panel was
therefore set up to collect new data describing the fabric of
sedimentary processes in order to build a new "spatialised" reference
model (*cf.* the "classical" model proposed by Bertran and Lenoble
2002). Please, do not hesitate to send referential data to feed these new models ! \bigbreak

## Contributors

The contributors panel aims at citing all contributors of this project
which is open to everyone interested in improving fabric analysis methods or
to contribute in building a new spatial model by sending reference data.
If any request or contribution, please contact the author of this
article. \bigbreak

# Util links

-   use Fabryka $\beta$ 3.0:
    <https://marchaeologist.shinyapps.io/fabryka/> \smallbreak
-   source codes: <https://github.com/marchaeologist/fabryka/tree/main>
    \smallbreak
-   github reprository doi of Fabryka $\beta$ 3.0: \smallbreak

# Acknowledgements

The writing of the script was greatly facilitated by the work published
by Shannon McPherron (2018). It includes some functions published by the
author, with a few modifications. This work benefited from discussions
with Jean-Pierre Texier, Arnaud Lenoble, Pascal Bertran and Dominique
Todisco on issues related to site formation processes. The reference data used to build the models of the classical method were collected by P. Bertran throughout his career. P. Bertran provided us with all these data in order to feed the application. The developpement of the application also benefited from the help of Rémi Lemoy and François and Victor Baleux on mathematical issues. A first version of the application was tested by
Emmanuel Discamps and Aurélien Royer. I would like to thank all these reaseachers for their advice and
support. \bigbreak

# Bibliography

The 'Bibliography' panel in the application contains the bibliography
cited in the application. This article contains the same bibliography
with some additional references. \bigbreak
