---
title: '**fabryka**'
subtitle: A web application for easily analysing & exploring the fabric of archaeological assemblages
output:
  pdf_document: default
header-includes:
- \usepackage{amsmath}
- \usepackage{lineno}
- \linenumbers
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{fabryka}
- \fancyhead[R]{R web application for fabric analysis}
- \fancyfoot[C]{\thepage}
- \usepackage{float}
- \floatplacement{figure}{H}
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: console
bibliography: references.bib
link-citations: true
---

\renewcommand{\[}{\begin{equation}}
\renewcommand{\]}{\end{equation}}
```{=html}
<style>
body {
text-align: justify}
</style>
```

\begin{figure} [H]
\centering
  \includegraphics[width=4cm]{figure/hex_Fabryka4.png}
\label{}
\end{figure}

Marc THOMAS ^1\*^ (<https://orcid.org/0000-0002-8160-1910>) \smallbreak

^1^ TRACES, UMR 5608 - Université Toulouse Jean Jaurès, 5 Allée Antonio
Machado, FR-31058 Toulouse Cedex 9 \smallbreak

^\*^ Corresponding author \smallbreak

E-mail: [marcthomas1\@hotmail.fr](mailto:marcthomas1@hotmail.fr){.email}
\bigbreak

**Key words** : R shiny web application, fabric, site formation
processes, spatial analysis, orientation and dip \bigbreak

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract

This article presents fabryka 1.1, an open access R shiny web
application dedicated to the fabric analysis of archaeological or
geological assemblages. The fabric is a statistic made of the
orientation and dip of elongated particles and archaeological objects.
It provides information about the nature of the geological processes
involved in the formation of the deposits and their impact on the
preservation of the archaeological assemblages. These statistical
analysis are not easily accessible to non-specialists and are
particularly time-consuming, sometimes requiring the use of several
software, some of which are chargeable. fabryka uses classical
analytical procedures (Benn, rose, Schmidt and Woodcock diagrams and statistical
tests on orientations) for analysing the fabrics of objects and includes
a spatial method. New statistical methods and new ways of exploration
are proposed in the application and this present article.\smallbreak

You can use fabryka by following this link
<https://marchaeologist.shinyapps.io/fabryka/> or by launching the
application in Rstudio thanks to its script. The fabryka application
files, made in R [@rcoreteam2025] by using the R package *shiny*
[@chang2024], and this article, written in rmarkdown [@allaire2014;
@xie2018; @xie2020], are both available online in a public github
repository (<https://github.com/marchaeologist/fabryka/tree/main>).
\pagebreak

# Introduction

Fabric is made of the orientation and dip of elongated objects
[@bertran1997; @bertran2002; @lenoble2004]. It is a statistic (*i.e.*,
it needs as much measurements as possible) that enables the
identification of sedimentary processes involved in the formation of the
deposits [@bertran1997; @bertran2002; @lenoble2004] and their impact on
archaeological assemblages [@isaac1967; @schick1986; @thomas2019].
Objects abandoned by prehistoric people can be considered as sedimentary
particles and are affected by the same processes [@schiffer1983]. As
geological particles are often not elongated enough, analysis of the
fabric of sedimentary deposits depends mainly on archaeological material
(bones and lithics). This makes it possible to identify the sedimentary
processes to which they were subjected and to discuss their degree of
spatial disorganization.

One of the first archaeological application probably dates to the
experiments carried out by the South African G. Isaac (1967) on the
shores of Lake Magadi in Kenya. G. Isaac wanted to explain the role
played by hydrological processes on the bone accumulations found in the
Great Rift Valley. In November 1964, he placed several concentrations of
experimental material near a river branch. A few months later, heavy
rains had reactivated the valley's rivers. In addition to the
displacement or even loss of all or part of the material, G. Isaac
revealed a reorientation that was mainly perpendicular and, to a lesser
extent, parallel to the orientation of flow, with a slight upstream dip.
Since G. Isaac's pioneering work, there has been an abundant literature
devoted to the analysis of fabrics [@benn1994; @lenoble2004;
@mcpherron2018] allowing this method to be increasingly performed on
archaeological sites.

Numerical methods have been developed to measure the orientation and dip
of objects in the field [@mcpherron2005artifact; @discamps2024] or from
field archives such as plans and photos [@de2013application;
@sanchez2016assessment].

Two main families of statistical method are commonly used to
characterize the fabric of a given assemblage [@bertran2002;
@mcpherron2018; @de2021new]. 

The first methods focus on the orientation
of the measurements only and aim to test the null hypothesis of a random
orientation from a uniform distribution through several statistical
tests [@krumbein1939; @curray1956; @rao1967large].

The second statistical method take into account object orientation and
dip. Three normalized eigenvalues (E1, E2 and E3) are calculated from
the cosine formed by the axis of the artifact with the three axes of
space [@watson1966]. @woodcock1977 and @benn1994 indices are used to
describe the fabric shape. Several authors *e.g.*, [@bertran1997;
@bertran2002; @lenoble2004] built models to compare the fabric of
archaeological assemblages to natural sedimentary processes and human
processes (*i.e.*, fabric of knapping spots). The null hypothesis of an
absence of perturbation is a fabric shape compatible with that of
experimental knapping spots. These methods are described in the
Mathematics section of this article.

Several software allow data processing, some of which are not freely
accessible, do not allow spatial exploration of the data or do not offer
interfaces to facilitate the task of non-programmer users
[@mcpherron2018].

The following sections explain all the selected approaches and our
proposals for processing the fabrics thanks to fabryka, from taking
measurements in the field, through the statistical methods and graphical
modes of representation used in the application (fig.
\ref{fig:figure_presentation}). \bigbreak

\begin{figure} [h]
\centering
  \includegraphics[width=16cm]{figure/presentation_applications.png}
  \caption{fabryka creation scheme and organisation of the article}
\label{fig:figure_presentation}
\end{figure}

# Field data acquisition

During excavation, the orientation and dip of elongated remains (*i.e.* a > 2 * b and with a > 2 cm) is
measured on a representative set of artefacts or on the whole
assemblage. The orientation (= bearing) of the lowest point is measured
in the range [0; 360] with North as the reference (*i.e.* orientation =
0°). The dip (= plunge) is measured in the range [0; 90] with horizontal
as a reference. The orientation [0; 360] can be converted into axial
data measured in the range [0; 180] by subtracting 180° from the
measurements in the range ]180; 360].

Traditionally, the orientation of the objects lowest point is measured
using a compass and the dip is measured using an inclinometer. The aim
is to measure the orientation and dip of the physical axis of elongation
of objects ignoring its shape. To do this, the measuring device must be
placed parallel to the general axis of elongation of the object.
Measuring the fabric from its imprint or visible surface is biased by
its shape, making impossible to assess its true dip. Since the
theodolite was introduced to excavations in the 1990s, the fabrics is
sometimes measured by recording the coordinates of the ends of objects [@mcpherron2005artifact].
In the latter case, the rules of trigonometry are applied to convert the
pair of coordinates into orientations and dips.

Recently, speleologist have created a digital inclinometer (Leica
DistoX310 modified to DistoX2; <https://paperless.bheeb.ch/>) which can
be used to measure fabrics [@discamps2024]. The device stores and
associates data with the order of measurement. When the device is tilted
downwards, the orientation of the lowest point is measured and, the
recorded dip is negative. When the device is tilted upwards, the
orientation of the highest point is recorded, and the dip is positive.
To obtain data that is suitable for statistical analysis, in the first
case the absolute value of the dip must be retained. In the second case
(*i.e.* positive dip), if the measured orientation is in the interval
[0; 180], 180° must be added to the orientation, and if the measured
orientation is in the interval ]180; 360], 180° must be subtracted from
the orientation.

When possible, preferring the use of a digital inclinometer avoids many
errors during measurement and data recording. On the other hand, for the
reasons given above, the use of a theodolite to measure fabrics is not
recommended as it results in inaccurate dip measurements. As well,
measuring orientations and dip in sections leads to sampling biases.

A sample corresponds to all the orientation and dip measurements taken
within a Lithostratigraphic Unit (LU) or a pedostratigraphic unit. These
units seem to be the most coherent entity which needs to be explored
because they are defined according to geological criterion, which a
priori guarantee a certain coherence in terms of the formation dynamics
we are seeking to approach. In fabryka, analysis are constrained within
samples unless the user decides otherwise in the spatial exploration
panel. \bigbreak

# Mathematics

## Classical method

Two main statistical family methods are used:

1.  The first considers the orientation only. The calculation of the
    Vector Magnitude $L$ (expressed as a percentage) by Curray's
    [-@curray1956] formula tests the hypothesis of a unimodal
    distribution of orientations. If the $p$ value returned by the
    Rayleigh test is less than a threshold of 0.05, the null hypothesis
    of a random orientation from a uniform distribution is rejected. The
    formula for calculating the Vector Magnitude $L$ is as follows:

Curray's formula to calculate $L$ is:

$$L = \frac{r*100}{n}$$ with:

$$r = \sqrt{\sum_{i=\alpha}^n \sin{(2\alpha)} + \sum_{i=\alpha}^n \cos{(2\alpha)}}$$
\smallbreak where $n$ is the number of measures and $\alpha$, is the
axial data (0 to 180°) \smallbreak

The Rayleigh test is performed thanks to the formula:

$$p = e^{(-(n*L^2))*10^-4}$$

The method consisting of doubling the value of the axial angles before
calculating the Vector Magnitude $L$ and the Rayleigh test, as proposed by
Krumbein [-@krumbein1939], tests the hypothesis of a bimodal
distribution with a period of 90°. The hypothesis of a 90° period
bimodal distribution of a sample of orientations is retained when $p$
\<0.05 [@curray1956]. The sensitivity of the Rayleigh test to the sample
size $n$ prevents samples from being compared with each other when $n$
varies [@curray1956; @bertran2002]. For example, when the Vector
Magnitude $L$ is 25%, the Rayleigh test gives $p-values$ below a critical
threshold of 0.05 for series with more than 50 measurements (fig. 2).

The Rao [-@rao1967large] spacing test assesses the uniformity of
circular data. It determines whether the data show significant
directionality (*i.e.* the null hypothesis is a uniform distribution).
It is widely used to detect multimodal directionality.

2.  The second considers the orientation and dip of the objects. The
    normalised eigenvalues $E1$, $E2$ and $E3$ are calculated from the
    cosine formed by the axis of the objects with the three axes of
    space [@watson1966]. 
    
The fabric shape can be described by calculating the ratios $r1$ = ln(E1/E2), $r2$ = ln(E2/E3) and $K$ = r1/r2 [@woodcock1977]. When 0 \< $K$ \< 1, the fabric is planar. When 1 \< $K$ \< $+\infty$, the fabric is linear. The parameter $C$ = ln(E1/E3) is a measure of the strength of the preferred orientation [@woodcock1977]. These indices can be plotted into a Woodcock diagram. 
    
The Benn indices [@benn1994] describe the fabric shape of a sample and can be plotted in a ternary diagram (*i.e.* Benn diagram). The elongation index EL is equal to $1-(E2/E1)$ and the isotropy index $IS$ is $E3/E1$. When $EL$ is high, the axes of the objects are grouped around an orientation. When $IS$ is high, the fabric is said to be isotropic meaning the axes of the objects are uniformly distributed in all directions. Finally, when the value $1-EL-IS$ is high, the fabric is said to be planar, with the axes of the objects grouped around a plane. \bigskip

```{r rayleigh, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Evolution of the p-value returned by the Rayleigh test as a function of the Vector Magnitude 'L' and the sample size 'n' \\label{rayleigh}", fig.height = 3.5, fig.pos="f"}

require(ggplot2)
L = c(0:60)

rayleigh <- function(L,n){
  p = exp(-(n*L^2)*10^-4)
  p_value <- p
  p_value
}

p_value_sim_30 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_30[i]<- rayleigh(i,30)
}

p_value_sim_40 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_40[i]<- rayleigh(i,40)
}

p_value_sim_50 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_50[i]<- rayleigh(i,50)
}

p_value_sim_100 <- rep(x = NA, times = length (L))
for (i in 1:61) {
  p_value_sim_100[i]<- rayleigh(i,100)
}

seuil_0.05 = c(rep(0.05, length(L)))

rayleigh_table <- data.frame(L, p_value_sim_30, p_value_sim_40, p_value_sim_50, p_value_sim_100,seuil_0.05)
  
ggplot(rayleigh_table, aes(x=L)) + 
    geom_line(aes(y = p_value_sim_100, colour = "n = 100"), size = 1) +
    geom_line(aes(y = p_value_sim_30, colour = "n =  30"), size= 1) +
    geom_line(aes(y = p_value_sim_40, colour = "n =  40"), size= 1) + 
    geom_line(aes(y = p_value_sim_50, colour = "n =  50"), size= 1) +
    geom_line(aes(y = seuil_0.05, colour = "threshold 0.05"), size= 0.5) +
    xlab("orientation rate 'L' (%)") +
    ylab("p-value") + 
    scale_color_manual(values=c("#0073C2", "#EFC000", "#868686", "#3B3B3B", "red")) +         
    scale_x_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60), expand = c(0, 0)) +
    scale_y_continuous(breaks = c(0, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.70, 0.80, 0.90, 1), expand = c(0, 0)) +
    theme_minimal() +
    theme(legend.position = "bottom") + 
    labs(colour = "")


```

## Spatialised method

To distinguish different statistical populations, a spatial method
proposed by McPherron [-@mcpherron2018] is used. Thanks to this method
it is possible to calculate the Benn indices and the Vector Magnitude
$L$ (formula (**3.**) modified from McPherron [-@mcpherron2018] R
script) for each remains, and its $n$ nearest neighbors in 3 dimensions
(McPherron [-@mcpherron2018] R script, modified, formula (**4.**)). The
analysis of the fabric as proposed by McPherron [-@mcpherron2018] has
the advantage of spatialising the measurements. It is possible to
localize the statistical signal of fabrics by varying the number of
measurements sampled to calculate the Benn indices. The choice of about
40 to 60 measurements (each remain and its 40 to 60 nearest neighbors)
to calculate the Benn indices allows local particularities to be
observed, and above all, it allows the Rayleigh test to be carried out
in its field of application (stability of the test between around 40 to
50 measurements after @bertran2002).

By avoiding the instability of the Rayleigh test, the rejection rate
$Tr$ of the null hypothesis provides a more reliable description of the
distribution of orientations in a sample. In addition, it allows
comparisons between samples despite varying sample size.

The rejection ratio $Tr$ is performed thanks to the formula:

$$Tr = \frac{(n~times~p.value < 0.05)* 100}{n}$$

# Graphical representation

Several methods of representation are used. Rose diagrams summaries the
orientation and dip measurements for each sample. Benn indices are
projected in a Benn diagram and measures (orientation and dip) are
projected into Schmidt diagrams. These figures are available for both
'classical' and 'spatialised' methods.

As submited by @mcpherron2018, a spatial investigation of the fabric is
proposed in a dedicated panel. @benn1994 indices calculated for each
spatial series of remains are projected within a Benn diagram in which
each pole (isotropic, linear, and planar) is colored respectively in
red, green and blue. The RGB color code is reused to materialize the
fabric of the remains on spatial projections (plan and section views) of
the material along the axes of the excavation grid. \bigbreak

# Interface and user interactivity

The next sections provide a description and a tutorial of fabryka. All
the figures in this section were produced using the example data set
provided in the application (*i.e.* "Case 3: Angles from DistoX2 with
coordinates).”

## Upload data

The simplest method to upload data in the application is to name the
column names directly, as shown in the example files. Enter your data in
a csv file, adding the columns 'id', ‘code’ and 'sample'. If the columns
don't exist, fill them in with the same letter. By doing so, you do not
need to select your variables in the main panel. The application
correctly reads the data file directly.

First, upload your .csv file in the left sidebar panel (fig.
\ref{fig:figure_upload_your_data}A). Pay attention to your decimal and
column delimiters (fig. \ref{fig:figure_upload_your_data}B and C).

fabryka supports three angle measurement methods:

-   angles taken with a compass and an inclinometer
-   angles taken with a DistoX2
-   angles taken with two shots at the total station.

Five cases (*i.e.* forms of data) are covered by the application (fig.
\ref{fig:figure_upload_your_data}D):

-   check 'Case 1: Only angles' if you have in your data set the columns
    'id', 'sample', 'code', 'orientation' and 'dip'.
-   check 'Case 2: Only angles from DistoX2' if you have in your data
    set the same columns that above but your data comes from a DistoX2
    and needs to be converted.
-   check 'Case 3: Angles from DistoX2 with coordinates' if you have in
    your data set the same columns that above (Case 2, data from
    DistoX2) with a single set of coordinates (x, y, z).
-   check 'Case 4: Angles and coordinates' if your data has the same
    form that above (Case 3) but is coming from a compass and an
    inclinometer.
-   check 'Case 5: Two shots data without angles' if you have the
    columns 'id', 'sample', 'code' and the angles have been measured
    with a total station (*i.e.* two shots data).

According to your method of angle measurement and the presence of
coordinates you must select the right case. Please refer to the example
files in the left side bar panel (fig.
\ref{fig:figure_upload_your_data}E). It contains a data set coming from
the Lower shelter of Le Moustier [@thomas2019; @texier2020]. The example files
directly contain the right column names. You can use them directly in
the application. Whatever the names of the columns in your file, you can
select the appropriate columns in the column selector (fig.
\ref{fig:figure_upload_your_data}F). Please keep only complete lines
(*i.e.* no empty modalities, so no line without a fabric measurement).
When a data table with new columns appears below the column selector,
you are ready to use the application.

If your data comes from a DistoX2, the columns 'orientation' (=
'bearing', *i.e.* 0 to 360°) and 'dip' (= 'plunge') are automatically
corrected in the new table. Whatever the form of your data set, two new
columns 'orientation_pi' (0 to 180° angles) and 'angle_double' (see
Krumbein 1939) appear. If the angles were measured using a total station
(*i.e.* two shots data or 2 by 2 coordinates of plotted pieces refits), the orientation and the dip are calculated. If only one set of coordinates is provided with angles, a second set of
coordinates is calculated and displayed in the new table (fig.
\ref{fig:figure_upload_your_data}G). You can download the new data by
clicking the 'copy', the 'csv' or the 'pdf' buttons. \bigbreak

## Classical method

The classical method (fig. \ref{fig:figure_classical_method}) can be
computed for all sources of data and absence/presence of coordinates
(*i.e.* all cases described above).

\begin{figure} [H]
\centering
  \includegraphics[width=16cm]{figure/upload_your_data.jpg}
  \caption{A. Browse your .csv file, B. choose your column separator, C. choose your decimal separator, D. check the right form of your dataset, E. to do so, have a look of example files, F. select the column needed in the application, G. a new dataset with new columns appears, your are ready to explore your data.}
\label{fig:figure_upload_your_data}
\end{figure}

This panel includes:

-   Benn diagrams with possible integration of @bertran2002 model (fig.
    \ref{fig:figure_classical_method}D)
-   Rose diagrams for axial data and dip (fig.
    \ref{fig:figure_classical_method}G)
-   Schmidt diagrams (fig. \ref{fig:figure_classical_method}H)
-   Woodcock diagrams
-   Summary data table with Benn indices, additional statistics, and
    orientation statistical tests.

In this last table (fig. \ref{fig:figure_classical_method}I):

-   $N$ stands for number of measures in the samples
-   $E1$, $E2$ and $E3$ are the eigenvalues [@watson1966]. They
    correspond to the sum of the cosines formed by the axis of the
    objects and the three axis of the space. They are used to compute
    the Benn indices
-   $IS$ and $EL$ are the Benn indices ratios standing
    respectively for, 'isotropy' with $IS = E3/E1$ and 'elongation' with
    $EL = 1-(E2/E1)$ [@benn1994].
-   $K$ and $C$ are the @woodcock1977 indices.
-   $L$ is the Vector Magnitude (strength of the preferred orientation)
    and $R.p$ is the p-value result of the Rayleigh test.
-   $L.double$ and $R.p.double$ are the same statistics that above for
    double angles [@krumbein1939].
-   $Rao_p$ is the p_value of the @rao1967large test.

You can control the data you plot and the data appearing in the figures
and the summary data table by changing the checkboxes filters ('filter
by sample' and 'filter by code') and the slider (*i.e.* minimum number
of measures) in the left sidebar panel (fig.
\ref{fig:figure_classical_method}A, B and C). 

\begin{figure} [H]
\centering
  \includegraphics[width=16cm]{figure/classical_method.jpg}
  \caption{A. filter by sample, B. if you want to filter by codes, uncheck 'no code distinction', C. manage the minimum sample size to perform analysis, D. buld your model thanks to Bertran and Lenoble (2002) data, E. and F. download figures by selecting a file format and pressing the 'download' button, G. Rose diagrams subpanel, H. Schmidt diagram subpanel, I. Summary table with additional statistics}
\label{fig:figure_classical_method}
\end{figure}

Certain geological processes can have a different impact on objects depending on their
physical properties (*e.g.* silex versus bones in a water flow as in the
example files provided where bones have a strong preferential
orientation and are much more 'linear' in the Benn diagram [@thomas2019; @texier2020]. The natural processes models that you can add to the Benn diagram are calculated using data mainly from
[@bertran1997; @bertran2002; @bertran2006; @lenoble2004]. You will find
the reference model data (with corresponding literature references)
which is used in the application in the 'Model reference data' subpanel.
In the rose diagram, axial data is considered (and not orientations) as
the Rayleigh test is perform on axial data [@curray1956] in the summary
table. You can download all the figures by selecting a file format
(.pdf, .jpeg and .png) and pressing the 'download' button (fig.
\ref{fig:figure_classical_method}E and F).

## Spatialised method

This panel includes (fig. \ref{fig:figure_spatialised_method}):

-   Benn diagrams with Benn indices computed with nearest neighbors
    points [@mcpherron2018]
-   Spatial projections of the points according to their fabric shape
-   Summary data tables with orientation statistical tests, one at the
    scale of each spatial series (each object and its $n$ nearest
    neighbors fabric = one row) and one at the scale of the sample (fig.
    \ref{fig:figure_spatialised_method}F and G).

A table similar to the one done in the Classical method is proposed and
contains the same statistics .

At the spatial series scale (*i.e.* each points and $n$ neighbors),
statistics are provided in a table (Spatial series data table subpanel)
with the Benn indices, the orientation rates $L$, $L.double$ and
$p-values$ returned by the Rayleigh tests. At the sample scale, mean and
standard deviations values of Benn indices, mean $L$ and its
corresponding Rayleigh tests means (mean values of all spatial series in
the sample) and rejections ratios $Tr$ (formula **4.**) are given in the
subpanel 'Summary table of samples'.

In the projections, the orientation of the objects can be materialized
by sticks (by unchecking the corresponding box in the left sidebar
panel, fig. \ref{fig:figure_spatialised_method}D) which color is
representative of the fabric of the series formed by the object and its
$n$ nearest neighbors. You can also add orthophoto in the background of
projections (fig. \ref{fig:figure_spatialised_method}E). In the left
sidebar panel, you can control the samples considered in the analysis by
selecting the sample explored or by distinguishing the code used (*e.g.*
flint versus bone, fig. \ref{fig:figure_spatialised_method}A and B). You
can also use the slider to change the number of neighbors searched for
around each point (fig. \ref{fig:figure_spatialised_method}C). Be
careful, this will alter the results obtained in the Benn diagram and
Rayleigh tests. \bigbreak

## Spatial exploration

This panel includes (fig. \ref{fig:figure_spatial_exploration}):

-   A spatial exploration of your data starting from a Benn diagram
-   A spatial exploration of your data starting from the spatial
    projections (fig. \ref{fig:figure_spatial_exploration}J)
-   Summary data tables with orientation statistical tests (fig.
    \ref{fig:figure_spatial_exploration}I and K).

In this panel, you can perform a spatial analysis of your data starting
from a Benn diagram or from projections. In both cases, you need to
select the data you want to explore more locally in a new analysis. From
the projections, you can select the points you want to explore directly.
In the Benn diagram, you need to select the lasso selector (fig.
\ref{fig:figure_spatial_exploration}D). Pay attention to the number of
points selected and check that the new sample contains enough points to
carry out the new analysis. If this is not the case, you can lower the
slider in the left sidebar panel to perform the analysis with fewer
neighbors (fig. \ref{fig:figure_spatial_exploration}E). You can download
the new samples you have selected with spatial statistics from the
corresponding subpanels (fig. \ref{fig:figure_spatial_exploration}I and
K). \bigbreak

\begin{figure} [h]
\centering
  \includegraphics[width=16cm]{figure/spatialised_method.jpg}
  \caption{A. filter by sample, B. if you want to filter by code, uncheck 'no code distinction', C. choose the n nearest neighbors to perform the statistics, D. make sticks in spatial projections, E. upload a georeferenced orthophoto and check 'include' orthophoto to display it, F. new dataset at the spatial series scale, G. summary table at the sample scale.}
\label{fig:figure_spatialised_method}
\end{figure}

## Tutorial

The 'Tutorial' panel contains a summary of the mathematics, statistics
and methods used in fabryka and a short tutorial directly accessible
from the application. \bigbreak

\begin{figure} [h]
\centering
  \includegraphics[width=16cm]{figure/spatial_exploration.jpg}
  \caption{A. if you want to filter by sample, uncheck 'No sample distinction' B. if you want to filter by code, uncheck 'no code distinction', C. choose the n nearest neighbors to perform the statistics (first Benn diagram), E. choose the n nearest neighbors to perform the statistics on the sampled data (second Benn diagram), F. make sticks in spatial projections, G. you can upload a georeferenced orthophoto and check 'include' orthophoto to display it, H. a new spatial exploration of the fabric is performed, I. new dataset at the selected spatial series scale, I. a similar spatial exploration is possible from the projections, K. new dataset at the selected spatial series scale.}
\label{fig:figure_spatial_exploration}
\end{figure}

## New reference models

The 'spatialised' method generates often more
linear fabrics in comparison with the 'classical' method. This can be
explained by the fact that objects that are close in space are more
likely to resemble each other than objects that are further in space. A difference between global fabric ("classical method", *i.e.*, one point per sample) and spatialised fabric 
(series of n neighbors) in Benn diagrams marks a spatial heterogeneity that needs to be explored.

Also, the structure of the point pattern is evocative of its fabric.
Thus, the dispersion of the points depends on the shape of the fabric.
Two extremes appear: the most isotropic samples are those with the most
dispersed point patterns (*i.e.* the standard deviations are high) while
the most linear samples are those with the least dispersed point
patterns. Planar samples highlight point patterns with intermediate
dispersion. The explanation of this phenomenon is intuitive: the more
similar the fabrics of objects are, the closer they are in the Benn
diagram and the more linear they are. To reflect this phenomenon, the
standard deviations of the Benn indices are calculated and aim to
describe the dispersion point patterns in the Benn diagrams.

Global planar or isotropic fabrics can sometimes mask local anisotropy.
Within certain samples, groups of objects form a dense cluster in the
point pattern. This is expressed either by a locally more linear fabric,
or by a planar fabric with a strong bimodal orientation.

For the reasons given above, we wish to create new models that
considered the particularities of the spatial method. A panel was
therefore set up to collect new data describing the fabric of
sedimentary processes in order to build a new "spatialised" reference
model (*cf.* the "classical" model proposed by @bertran2002). This work
is in progress. This panel will be updated in the next version of the
application.

\bigbreak

## Contributors

The contributors panel aims at citing all contributors of this project. \bigbreak

# Useful links

-   use fabryka online: <https://marchaeologist.shinyapps.io/fabryka/>
    \smallbreak
-   If you are preparing a mission in a field without Internet access:

Install [R](https://www.r-project.org) and [Rstudio
Desktop](https://posit.co/download/rstudio-desktop/).

Clone the fabryka repository from GitHub:
<https://github.com/marchaeologist/fabryka> or download the code in
"Code" and "Download ZIP" at the same address. This will give you the
latest version of the source code. The application can be launched by
opening the app.R file in the main folder and by running the following
commands:

```{r, eval=FALSE}

pkgload::load_all(".")
fabryka()

```

If you are using the application for the first time, Rstudio will install several packages used in the application. Lists of all these packages and dependencies are available in a DESCRIPTION file and a DEPENDECIES file in the github repository.

-   source code: <https://github.com/marchaeologist/fabryka/tree/main> \smallbreak
-   Report bugs or any request: <https://github.com/marchaeologist/fabryka/issues> \smallbreak
-   github reprository doi of fabryka:
    <https://doi.org/10.5281/zenodo.15207419> \smallbreak
-   Zenodo repository of this article:
    <https://doi.org/10.5281/zenodo.15879462> \bigbreak

# Acknowledgements

The writing of the script was greatly facilitated by the work published
by Shannon McPherron (2018). It includes some functions published by the
author, with a few modifications. This work benefited from discussions
with Jean-Pierre Texier, Arnaud Lenoble, Pascal Bertran and Dominique
Todisco on issues related to site formation processes. The reference
data used to build the models of the classical method were collected by
P. Bertran throughout his career. P. Bertran provided us with all these
data in order to feed the application. The development of the
application also benefited from the help of Rémi Lemoy and François and
Victor Baleux on mathematical issues. A first version of the application
was tested by Emmanuel Discamps and Aurélien Royer. This project is
supported by the "Service Valorisarion de la Recherche" of the
University Toulouse Jean Jaurès. Finally, the constructive comments of
Pascal Bertran, recommender of this article, and, of Frédéric
Santos, Nicolas Frerebeau and Alfonso Benito-Calvo, reviewers of this
article, considerably improved the manuscript and the web application. I
would like to thank all these reaseachers for their advice. \bigbreak

# Conflict of interest disclosure

The authors declare that they comply with the PCI rule of having no financial conflicts of interest in relation to the content of the article. \bigbreak

# Bibliography

The 'Bibliography' panel in the application contains the bibliography
cited in the application. This article contains the same bibliography
with some additional references. \bigbreak
